---
title: 'A time series analysis about the crime in Puerto Rico from the year 2007 to 2014'
author: "Juan De Dios Santos"
date: "December 5, 2015"
output: html_document
---

## Overview
***
Crime Pattern Detection is one of the many applications of data mining; it
deals with the study, analysis and understanding of crimes through discovery of pattern
or other significative insight in data. As you might think, 
detecting a pattern of criminal activity, or even predicting them is something that might seem 
taken from a movie. However, it is a real application of data mining, and there 
are many research groups working on that.

In this report I will introduce the concept of time series analysis using the R base
package, to inspect the assassination history of Puerto Rico from 2007 to 2014.

## The data
***

This report's dataset features the total amount of murders perpetrated in Puerto
Rico during the period 2007 - 2014. This data is divided by months and years. 

Let's take a look at it.


```{r}
pr_crime_table <- read.csv("~/Desktop/pr_crime_table.csv", row.names = 1)
print(pr_crime_table)
```

The first row of the dataset (the header) are the months, and the first column (row names)
represent the year. For example, on January 2007 there were 66 registered assassinations.

The data was obtained from the [_Instituto de Estadisticas de Puerto Rico_](http://www.estadisticas.gobierno.pr/iepr/)
(Statistics Institute of Puerto Rico).

## The analysis
***

Now that we have seen how the data is, let's commence the analysis by building
the time series model. To build the model we need a vector of the values to
be analyzed (the frequency of crimes), and the time interval, seen as `frequency`
on the code, which is 12 since we are using monthly data, and lastly we need
to state the start data (January 2007).

```{r, echo=FALSE}
# Transform the table into a vector; ts should just take a vector, it performs
# the split into times by itself
ts.model <- ts(c(t(pr_crime_table)), frequency = 12, start = c(2007, 1))
```

Once the model is built, the next step is to decompose it into several components.
These components are: seasonal, trend and irregular. In the next section, I will
explain these concepts in detail.

To decompose the data into the previously mentioned elements, I applied a technique
named moving averages - a calculation that shows the average value of
a subset over a period of time.

```{r}
decomposition.ma <- decompose(ts.model)
names(decomposition.ma)
```

The previous output shows the components of this `decomposed` object. As it was
previously mentioned, our objective was to decompose the time series into a 
seasonal, trend and irregular (named `random` in the object) component.

### The seasonal trend
***

The seasonal trend is the common trend across all seasons (months).

```{r}
plot(decomposition.ma$figure, type="b", xaxt="n", xlab="")
months.names <- months(ISOdate(2007,1:12,1))
axis(1, at=1:12, labels=months.names, las=2)
```


The past plot is the seasonal trend of the data. What you see in the figure is the
generalized trend of the crime activity during a year, based on seven years of crime.
The x-axis represent the months of the year, and the y-axis shows the difference between 
the number of crimes, and the global average number of crimes. For example, 
suppose that the average number of murders in a month is 20, then the average during March 
would be 15 (20 - 15), because the March average is -5.

The plot shows that during the period February - April, the amount of crimes decreases
until it reaches a global minima at -13.5. Then, on the following month, May, the
average number of murders increase by 19. Thus, we have the less active month,
and the most active back to back. From April until the end of the year, the crime
activity does not vary that much. The following table shows the exact values of
the plot.

```{r}
print(decomposition.ma$figure)
```

### The trend
***

The trend decomposition is the trend of the murders across the complete dataset.

```{r}
plot(decomposition.ma$trend, xlab = 'Year', ylab = 'Number of assassinations',
     main = 'Trend of assassinations in Puerto Rico during 2007 - 2014')
```

We can see on this plot that the assassinations increases until it reaches a maximum
point during mid-2011, meaning that 2011 was the year with the most assassinations.
After that year the number decline. 

The following table presents the total murders per year. If we compare those numbers
with the plot, we can notice that it follows the pattern presented on the previous
image. From 2007 until 2011, the frequency of murders
increment up to 2011 (this is the peak on the plot). Afterwards, from 2011 until 2014 the recurrence of homicides slows down.

```{r}
apply(pr_crime_table, 1, function(x) sum(x))
```

## Decomposition plot

For the last section of the report, I will present a summary of the decomposition
using a visualization that has the previous two figures, plus a plot of the entire
original dataset, and the random or remainder decomposition. Each value of the
remainder component is the difference between the original value from the seasonal
plus trend fit.

```{r}
plot(decomposition.ma)
```

The first plot is the original values. If we compare it to the second one, the trend,
we can notice that it follows the same structure, a peak at mid-2011 and
smaller values at the extremes. The third plot, the seasonal decomposition, is the
same seasonal trend plot presented before, but repeated for all the years. Finally,
the random plot is the remainder decomposition.

### Closure
***

In this report, I presented the basics of a time series analysis using the base R
package, and an application of it using crime data from Puerto Rico. The results
found indicates that there is a trend of crimes that takes place during a year;
during the first three months the homicide activity declines, followed by an 
increase in April, where it hits its maximum value, and subsequently it decreases 
and continues an almost similar behaviour until the end of the year. Moreover, 
the general trend of the data is that from 2008 up until 2011, the number of crimes
increased at a slow pace, and after mid-2011 it decreased.
